# Plotting

```{r, echo=F}
library(reticulate)
```

## Plotting basics

For the R and Python sections here, either pick one of the languages, or do both.

### R

#### Greek symbols

Take the following plot:

```{r}
plot(sin)
```

Change it so that the `x` axis label is $\sigma$ and the `y` axis label is $\tau$.
Those are the Greek letters "sigma" and "tau", respectively.

Hint: look up the R function `expression`.

```{r}
plot(sin, xlab=expression(sigma), ylab=expression(tau))
```

Take the same plot, and change the `x` axis label to $\sigma^2_s$.

#### Color and line style (base graphics)

For the same plot as above, change the plot so that the line is red, "dot-dash", and 3 times the default line thickness.

#### Axis labels from data

Take the graph we started with, and assume there's a variable `y = 7`.
Add a title to the graph that says $\Gamma = 7$.
(That's the Greek letter "gamma".)
Do it using the variable! 
Don't "hard-code" in the value!

Hint: the function you need here is `bquote`.
Extra hint: this is one of the more obnoxious corners of R.


#### Annotations from data

Take the graph we started with, and assume there's a variable `y = 7`.
Add a text element at coordinates `x=0.1, y=0.4` to the graph that says $\Gamma = 7$.

### Python

Do the same manipulations described above for the R plots, but using this graph:

Hint: all of the Greek letter stuff can be accomplished with string literals and $\LaTeX$ syntax.
(This is an area where Python is far easier than R.)

```{python}
import matplotlib.pyplot as plt

f, ax = plt.subplots()
x = [i for i in range(10)]
y = [i**2 for i in x]
ax.plot(x, y)
ax.set_xlabel('x')
ax.set_ylabel('y')
```

## Questions from the group

#### Narges

Heat maps with marginals in Python.

```{python}
import seaborn as sns

df = sns.load_dataset('iris')
# sns.jointplot(x=df["sepal_length"], y=df["sepal_width"], kind='kde')
# sns.jointplot(x=df["sepal_length"], y=df["sepal_width"], kind='scatter', s=200, color='m', edgecolor="skyblue", linewidth=2)
# sns.set(style="white", color_codes=True)
sns.jointplot(x=df["sepal_length"], y=df["sepal_width"], kind='kde', palette="crest", color="skyblue")
plt.show();
```


